Climate variability indices for ecological and crop models in R: the `climatrends` package

### 2020-05-25 inst/paper/paper.Rmd
paper_cbeans

---
title: 'Climate variability indices for ecological and crop models in R: the `climatrends` package'
tags:
- climate data
- climatology
- earth science
- evapotranspiration
- precipitation data
- R
- reproducibility
- weather data
authors:
  - name: Kauê de Sousa
    orcid: 0000-0002-7571-7845
    affiliation: "1, 2"
  - name: Jacob van Etten
    orcid: 0000-0001-7554-2558
    affiliation: 2
  - name: Svein Ø. Solberg
    orcid: 0000-0002-4491-4483
    affiliation: 1
affiliations:
  - name: Department of Agricultural Sciences, Inland Norway University of Applied Sciences, Hamar, Norway
    index: 1
  - name: Bioversity International, Rome, Italy
    index: 2
citation_author: de Sousa et. al.
date: "`r format(Sys.time(), '%d %B %Y')`"
year: 2020
bibliography: paper.bib
output: pdf_document
---

```{r get data, include=FALSE}
library("knitr")
library("kableExtra")
library("magrittr")
funtbl <- read.csv("available_functions.csv")
```

@@

**climatrends** aims to provide the R [@RCoreTeam] toolkit to compute extreme precipitation and temperature indices that serves as input for climate and crop models [@vanEtten2019; @Kehel2016], trends in climate change [@Aguilar2005; @deSousa2018] and applied ecology [@Prentice1992; @YLiu2018].

@@

```{r characteristics, echo=FALSE, message=TRUE}
kable(funtbl, 
      row.names = FALSE,
      align = "l", 
      caption = "Main functions available in climatrends.") %>% 
  kable_styling(latex_options = "scale_down") %>% 
  column_spec(2, width = "1in") %>% 
  column_spec(3, width = "1.7in") %>% 
  column_spec(4, width = "1.7in")
```

@@

Replicate part of the analysis in van Etten et al [-@vanEtten2019] with the beans data to show how we can use this package to capture the influence of climate variability on crop performance. The idea is to show the same PlackettLuce Tree.

@@

>>>>

---
title: 'Climate variability indices for ecological and crop models in R: the `climatrends` package'
tags:
- climate data
- climatology
- earth science
- evapotranspiration
- precipitation data
- R
- reproducibility
- weather data
authors:
  - name: Kauê de Sousa
    orcid: 0000-0002-7571-7845
    affiliation: "1, 2"
  - name: Jacob van Etten
    orcid: 0000-0001-7554-2558
    affiliation: 2
  - name: Svein Ø. Solberg
    orcid: 0000-0002-4491-4483
    affiliation: 1
affiliations:
  - name: Department of Agricultural Sciences, Inland Norway University of Applied Sciences, Hamar, Norway
    index: 1
  - name: Bioversity International, Rome, Italy
    index: 2
citation_author: de Sousa et. al.
date: "`r format(Sys.time(), '%d %B %Y')`"
year: 2020
bibliography: paper.bib
output:
  pdf_document:
    keep_tex: true
header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

```{r get data, include=FALSE}
library("knitr")
library("kableExtra")
library("tidyverse")
library("climatrends")
library("PlackettLuce")
funtbl <- read.csv("available_functions.csv")
load("cbean.rda")
```

@@

`climatrends` aims to provide the R [@RCoreTeam] toolkit to compute extreme precipitation and temperature indices that serves as input for climate and crop models [@vanEtten2019; @Kehel2016], trends in climate change [@Aguilar2005; @deSousa2018] and applied ecology [@Prentice1992; @YLiu2018].

@@

@@

During five growing seasons (from 2015 to 2016 in Nicaragua), van Etten et al [-@vanEtten2019] conducted a crowdsourcing citizen-science trial testing 11 common beans varieties (*Phaseolus vulgaris* L.) in 842 farmer-managed experimental plots. Sets of three varieties were allocated randomly to farms as incomplete blocks, maintaining spatial balance by assigning roughly equal frequencies of the varieties to each area. For the analysis the authors applied the Plackett–Luce model, which estimates for each variety the probability that it wins, beating all other varieties in the set [@Turner2020]. The research showed that this approach can register known specific effects of climate variation on crop varietal performance. An earlier version of `climatrends` was used in this research to capture the seasonal climate variation, here we replicate part of this analysis in the matter that concerns the calculation and application of the climate indices.

Is important to mention that this analysis shaped the implementation of the methods `temperature.array()` and `rainfall.matrix()`.

```{r ggd_bean, message=FALSE, echo=TRUE, eval=FALSE}
library("climatrends")
library("PlackettLuce")
library("tidyverse")

# compute the number of days required to accumulate
# gdd from planting date to maturity
gdd <- GDD(modis, 
           day.one = cbean$planting_date, 
           degree.days = 900, 
           return.as = "ndays")

# add gdd to the cbean data and take the average 
# of gdd per season
cbean %<>%  
  mutate(gdd = gdd$gdd) %>% 
  group_by(season) %>% 
  mutate(gdds = as.integer(mean(gdd)))
```

```{r pltree_bean, message=FALSE, echo=TRUE, eval=FALSE}
# compute the temperature indices from planting date to the 
# number of days required to accumulate the gdd in each season
temp <- temperature(modis, 
                    day.one = cbean$planting_date, 
                    span = cbean$gdds)

# combine the indices with the main data
cbean <- cbind(cbean, temp)

# fit a Plackett-Luce tree
plt <- pltree(G ~ maxNT, data = cbean, minsize = 200)
```

@@

<div id="refs"></div>

\pagebreak

# Figures and tables

```{r table1, echo=FALSE, message=TRUE}
kable(funtbl, 
      row.names = FALSE,
      align = "l", 
      caption = "Main functions available in climatrends.") %>% 
  kable_styling(latex_options = "scale_down") %>% 
  column_spec(2, width = "1in") %>% 
  column_spec(4, width = "1.7in")
```

\pagebreak

```{r fig_cbean, fig.cap="Fig. 1. Application of climatrends functions to support the analysis of a citizen-science data testing 11 common bean varieties in Nicaragua. (A) Days required to reach 900 growing-degree days from planting date calculated using the function GDD(). (B) Maximum night temperature (°C) distributed across seasons computed using the function temperature(). (C) Plackett-Luce Tree showing the probability of one common bean variety has to win against the others (axys X) in two different nodes splitted with the maximum night temperature (°C).", out.width = '80%', echo=FALSE}
knitr::include_graphics("cbean.png")
```
